\documentclass[sigconf]{acmart}

\usepackage{booktabs} % For formal tables
\usepackage[utf8]{inputenc}
\usepackage{listings}
\usepackage[inline]{enumitem}
\usepackage{FiraMono}

\input{misc/functions}

\lstset{%
  language=Haskell,
  basicstyle=\scriptsize\ttfamily,
  literate=
  {->}{{$\rightarrow$}}1 {=>}{{$\Rightarrow$}}1 {>>=}{{$\bindOp$}}2 {>>}{{$\bindIgnOp$}}1
  {<-}{{$\leftarrow$}}1
}

% Copyright
%\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

\begin{document}
\title{Why Data is the Better Monad}
\subtitle{Using Freedom to Great Effect}

\author{Justus Adam}
\affiliation{}
\email{justus.adam@tu-dresden.de}


\begin{abstract}
Side effects are the sore spot for pure languages. The mathematical beauty of
referential transparency is broken to allow outside interactions that are
unpredictable and hard to reason about. Interactions such as network and
filesystem access for example, that depend on invariants in the context they are
executed in and may unexpectedly fail, but are necessary to write useful
programs.

Fortunately, with the Monad, an abstraction was discovered that allowed safe and
transparent, sequential interaction with an arbitrary ambient environment.

Yet it too is flawed. Versatile though the Monad is, composing effects in a
reusable fashion proved difficult and has performance implications which
increase in severity as the granularity of effects gets finer.

With the free monad we now finally not only obtain a way to compose arbitrary
effects together in an efficient way but can also easily implement effects via
transformation into other effects and even implement the same effect in
different ways depending on needs and constraints of its application.

This paper provides an overview of the origins of the Monad, how its
shortcomings lead to the invention of transformers, then the free monad, and
eventually the freer monad. How free and freer monads work and how they can be
used to realise our collective dream of pure, efficient, composable and
versatile ambient interactions.
\end{abstract}

\maketitle

\input{Chapters/Introduction}
\input{Chapters/Monads}
\input{Chapters/Transformers}
\input{Chapters/Free}
\input{Chapters/BetterOpenUnion}
\input{Chapters/BindPerformance}
\input{Chapters/Freer}
\input{Chapters/AlternativeSystems}
\input{Chapters/Conclusion}

\bibliographystyle{ACM-Reference-Format}
\bibliography{bibliography}

\end{document}
